Staged Abstract Interpreters

### 2019-09-04 paper/60_cases.tex
revision

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{-2em}
\subsection{Control-flow Analysis} \label{cfa}

In this section, to demonstrate the SAI approach is widely applicable, we extend
and refactor the staged abstract interpreter we presented in Section \ref{sai}
to various control-flow analyses (CFA). We first discuss calling-context
sensitive analysis for staged analyzers by tweaking allocation strategy; then we
will see how to implement store-widening with staged monads, which is a common
optimization in CFA; finally, we integrate abstract garbage collection with
staging. All of them are in fact orthogonal to staging.

@@

The abstract addresses we introduced in Section \ref{sai} are simply identical to
the variable names, which is a context-insensitive schema.
Through tweaking the address allocation strategy, we can achieve various
context-sensitive analyses \cite{DBLP:conf/icfp/Gilray0M16}. In this section, we
demonstrate a calling-context-sensitive analysis, i.e., $k$-CFA-like analysis
\cite{DBLP:journals/jfp/HornM12}, in our staged abstract interpreter.
\begin{lstlisting}
  type Time = List[Expr]
  case class KCFAAddr(x: Ident, time: Time) extends Addr
\end{lstlisting}

The timestamp is essentially a finite list of expressions that tracks the $k$
recent calling contexts. The definition of abstract addresses @KCFAAddr@ is also
changed to contain an identifier and the time when it get allocated, meaning that
this address points to some values under such calling context. If $k$ is 0, we
obtain a monovariant analysis as demonstrated before; if $k > 0$, we obtain a
family of analysis with increasing precision. Note that when $k = 0$, the
address space is statically determined by the set of variables appeared in the
program, and we may exploit it to optimize the generated code. But when $k > 0$,
the address generation happens at the next stage, i.e., analysis time.
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepStateT[
                   RepStateT[RepSetReaderStateM[Cache, Cache, ?], Time, ?], //add Time with a RepStateT
                 Store, ?], Env, T]
  type Result = (Rep[Set[(Value, Store, Time)]], Rep[Cache])
\end{lstlisting}

We then integrate the timestamp into our monad stack by adding a staged
@StateT@ monad. The result type is passingly get updated to contain a set of
tuples, where the time is added.
\begin{lstlisting}
  def tick(e: Expr): AnsM[Unit] = for {
    τ <- get_time; u <- liftM(StateTMonad.put((e::τ).take($k$)))
  } yield u
  def alloc(x: String): AnsM[Addr] = for {
    τ <- get_time
  } yield unit(KCFAAddr(x, τ)) // we use unit to turn a current-stage constant to next-stage value
\end{lstlisting}

Every time when we call the @eval@ function, we refresh the timestamp by calling
the @tick@ function, which updates the timestamp in the state monad. The @tick@
function is implemented as prepending the current expression being evaluated @e@
to the existing calling context, and then taking the first $k$ elements from the list.
The type @Config@ used in caching is also changed to included the timestamp.

@@

A commonly used technique to improve the running time in control-flow analysis
is store-widening. As shown by \cite{Darais:2015:GTM:2814270.2814308,
DBLP:journals/pacmpl/DaraisLNH17}, store-widening analysis can be easily
implemented in monadic abstract interpreter by swapping the @StateT@ (for store)
and the @SetT@ (for nondeterminism). In this section, we show this optimization
is orthogonal to staging and we just need to swap the same \textit{staged}
version of monads. After the adjustment, we have the following monad stack:
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepSetT[RepStateT[
                   RepReaderT[RepStateT[IdM, Cache, ?], Cache, ?], Store, ?], ?], Env, T]
\end{lstlisting}

@@

Abstract garbage collection is a technique to reclaim unreachable addresses in
the store while doing the abstract interpretation
\cite{Might:2006:IFA:1159803.1159807}. \citet{DBLP:journals/pacmpl/DaraisLNH17}
shows that, for big-step monadic abstract interpreter, we may add a @Reader@
monad to track a set of root addresses and compute the reachable addresses every
time when we obtain a value. Here we adopt it with staged monads:
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepReaderT[
                   RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?],
                 Set[Addr], ?], Env, T] //add a reader monad for Set[Addr]
\end{lstlisting}

The idea of abstract garbage collection is shown in Figure \ref{fixgc}. We modify the @fix@
function with caching (from Section \ref{sai}) to @fix_gc@, but the caching part
is omitted. Several auxiliary functions are used: @ask_root@ retrieves the
current root addresses from the monad stack; @root@ compute the reachable
addresses from a value; and @gc@ cleans up the store according to the reachable
addresses and returns a new store. For the generic interpreter, we also need to
compute the root addresses for compound expressions, readers may refer to
\cite{DBLP:journals/pacmpl/DaraisLNH17}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Control-flow Analysis} \label{cfa}

In this section, to demonstrate the SAI approach is widely applicable, we
extend and refactor the staged abstract interpreter we presented in Section
\ref{sai} to various control-flow analyses (CFA). We first discuss
calling-context sensitive analysis for staged analyzers by tweaking the
allocation strategy. Then we describe how to implement store-widening with
staged monads, which is a common optimization in CFA. Finally, we integrate
abstract garbage collection with staging. All of them are in fact orthogonal to
staging.

@@

The abstract addresses we introduced in Section \ref{sai} are simply identical
to the variable names, which is a context-insensitive schema.  Through tweaking
the address allocation strategy, we can achieve various context-sensitive
analyses \cite{DBLP:conf/icfp/Gilray0M16}. In this section, we demonstrate a
calling-context-sensitive analysis, i.e., $k$-CFA-like analysis
\cite{DBLP:journals/jfp/HornM12}, can be implemented our staged abstract
interpreter.
\begin{lstlisting}
  type Time = List[Expr]
  case class KCFAAddr(x: Ident, time: Time) extends Addr
\end{lstlisting}

The timestamp is essentially a finite list of expressions that tracks the $k$
recent calling contexts. The definition of abstract addresses @KCFAAddr@ is also
changed to include the identifier and the time when it get allocated, meaning that
this address points to some values under such calling context. If $k$ is 0, we
obtain a monovariant analysis as demonstrated before; if $k > 0$, we obtain a
family of analysis with increasing precision. Note that when $k = 0$, the
address space is statically determined by the set of variables appeared in the
program, and we may exploit it to optimize the generated code. But when $k > 0$,
the address generation happens at the next stage, i.e., analysis time.
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepStateT[
                   RepStateT[RepSetReaderStateM[Cache, Cache, ?], Time, ?], //add Time with a RepStateT
                 Store, ?], Env, T]
  type Result = (Rep[Set[(Value, Store, Time)]], Rep[Cache])
\end{lstlisting}

We then integrate the timestamp into our monad stack by adding a staged
@StateT@ monad. The result type is accordingly updated to contain a set of
tuples, where the time is added.
\begin{lstlisting}
  def tick(e: Expr): AnsM[Unit] = for {
    τ <- get_time; u <- liftM(StateTMonad.put((e::τ).take($k$)))
  } yield u
  def alloc(x: String): AnsM[Addr] = for {
    τ <- get_time
  } yield unit(KCFAAddr(x, τ)) // we use unit to turn a current-stage constant to next-stage value
\end{lstlisting}

Every time when we call the @eval@ function, we refresh the timestamp by
calling the @tick@ function, which updates the timestamp in the state monad.
The @tick@ function is implemented as prepending the current control expression
being evaluated (@e@) to the existing calling context, and then taking the
first $k$ elements from the list.  The type @Config@ used in caching is also
changed to include timestamps.

@@

A commonly used technique to improve the running time in control-flow analysis
is store-widening. As shown by \cite{Darais:2015:GTM:2814270.2814308,
DBLP:journals/pacmpl/DaraisLNH17}, store-widening analysis can be easily
implemented in monadic abstract interpreter by swapping the @StateT@ (for store)
and the @SetT@ (for nondeterminism). In this section, we show this optimization
is orthogonal to staging and we just need to similarly swap the same \textit{staged}
version of monads. After the adjustment, we have the following monad stack:
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepSetT[RepStateT[
                   RepReaderT[RepStateT[IdM, Cache, ?], Cache, ?], Store, ?], ?], Env, T]
\end{lstlisting}

@@

Abstract garbage collection is a technique to reclaim unreachable addresses in
the store while doing the abstract interpretation
\cite{Might:2006:IFA:1159803.1159807}. \citet{DBLP:journals/pacmpl/DaraisLNH17}
shows that, for big-step monadic abstract interpreter, we may add a @Reader@
monad to track a set of root addresses and compute the reachable addresses every
time when we obtain a value. Here, we adopt the idea with staged monads:
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepReaderT[
                   RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?],
                 Set[Addr], ?], Env, T] //add a reader monad for Set[Addr]
\end{lstlisting}

The idea of abstract garbage collection is shown in Figure \ref{fixgc}. We
update the @fix@ function with caching (from Section \ref{sai}) to @fix_gc@,
but the caching part is omitted in the presented code. Several auxiliary
functions are used: @ask_root@ retrieves the current root addresses from the
monad stack; @root@ computes the reachable addresses from a value; and @gc@
performs garbage collection according to the reachable addresses and returns a new
store. For the generic interpreter, we also need to compute the root addresses
for compound expressions, readers may refer to
\cite{DBLP:journals/pacmpl/DaraisLNH17}.
